Experience Report: Haskell in Computational Biology

### 2012-03-10 mrfy_experience_report.tex
hammering on the HMM section complete

\subsection{Perspicuous Haskell}

Given a hidden Markov model, 
an~established software package called HMMER (``hammer'') 
can compute the probability
that a new protein shares structure and
evolutionary history with the proteins used to train the model.
The computation finds the most likely path through the hidden Markov model.
To~make best use of floating-point arithmetic, the software computes
the \emph{logarithm} of the probability of each path, by summing
the logs of the 
probabilities on the states and edges of the path \cite{Viterbi:1967hq}.
(The logarithm function is monotonic, so the path that maximizes the
log of the probability is the most likely path.)
The~specification is shown on the left-hand side of \figref{viterbi}.
The probability $V_j^M(i)$ represents the probability of the most
likely path of the first~$i$ amino acids in the query sequence,
terminating with placement of amino acid $x_i$ in state~$M_j$.
The probabilities $V_j^D(i)$ and $V_j^D(i)$ are similar.

@@

Another of us was curious to try to code Viterbi's equations
directly as recursive functions.
But like the classic recursive Fibonacci function, Viterbi's functions,
when implemented \naive ly,
perform extraordinarily badly.
But just like the Fibonacci function, Viterbi's functions can be
\emph{memoized}.
To~show the memoized code, we have to confess to a small
transformation of Viterbi's equations.
The logarithms on the left of \figref{viterbi} are computed when the
hidden Markov model is constructed, and the resulting numbers are
stored in a file.
Moreover, to make the file more pleasant to read, the logarithms are
negated.
(The logarithm of a probability is typically negative, and negating
the logarithm eliminates hordes of minus signs from the file.)
The negated log probability is called a \texttt{Score}.
We~therefore implement a transformed version of Viterbi's equations,
shown on the right-hand side of \figref{viterbi}, which minimize the
negated log probability for each combination of node~$j$, amino
acid~$x_i$, and state $M_j$, $I_j$, or $D_j$.
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{smallverbatim}
viterbi :: State -> Int -> Int -> (Score, StatePath)
viterbi Mat j i = ( eProb j i + minimum $ [ trans m_m Mat
                                          , trans i_m Ins
                                          , trans d_m Del 
                                          ]
                  , Mat : path )
  where (prevV, path) = viterbi prev (j-1) (i-1)
        trans t prev = prevV + tProb hmm (j-1) t
\end{smallverbatim}
ALTERNATE VERSION FOLLOWS
\begin{smallverbatim}
viterbi :: WHAT IS MY TYPE????
viterbi s n o
  | s == mat = 
       let trans t prev = 
        (scr + tProb (n-1) t + (eProb n o), mat:p)
        where (scr, p) = viterbi prev (n-1) (o-1)
       in minimum $ [
           trans m_m mat,
           trans i_m ins,
           trans d_m del
           ]
\end{smallverbatim}
% Similarly, the insertion equation is:
% \begin{verbatim}                    
%   -- consume an observation but not a node
%   | s == ins = let trans t prev =
%                    (score + tProb n t + (eProb n o), ins:path)
%                     where (score, path) = viterbi prev n (o - 1)
%                 in minimum [
%                     trans m_i mat,
%                     trans i_i ins
%                     ]
% \end{verbatim}
% Finally, the deletion equation is:
% \begin{verbatim}                    
%   -- consume a node but not an observation
%   | s == del = let trans t prev = 
%                       (score + tProb (n-1) t, del:path)
%                       where (score, path) = viterbi prev (n - 1) o
%                 in minimum [
%                     trans m_d mat,
%                     trans d_d del
%                     ]
% \end{verbatim}

In this second approach, we used the
\texttt{Data.Memocombinators} library to provide the memoization container 
(again, a three-dimensional array).
This combinator approach was easy to apply to the code after the fact:
\begin{smallverbatim}
viterbi state node obs = 
  Memo.memo3 (Memo.arrayRange (mat, end)) 
             (Memo.arrayRange (0, numNodes))
             (Memo.arrayRange (0, seqlen)) 
             viterbi' state node obs
\end{smallverbatim}
We determined that the run-time 
performances of both approaches were virtually indistinguishable, and we also 
noted that the top-down version quite faithfully resembled the recurrence 
relations found in a textbook~\cite{Durbin:1998wz} or the HMMER literature~\cite{Eddy:1998ut} 
(see Equation \ref{viterbi_eqn}).
We chose to keep the top-down version, which 
later proved to be wise when we had to track down a bug in one of our base 
cases.
This bug was observed as a missing state in the alignment our program 
displayed, and we quickly discovered that one of our base cases was missing 
this state: it was returning an empty list as the base of the path, rather than 
a list containing a single node.
In this, we were grateful for the resemblance 
between the mathematical description of the algorithm and the top-down 
dynamic-programming approach in Haskell, which resulted in perspicuous code.

Standard implementations of the Viterbi algorithm, such as in HMMER, spend much of their code
managing the dynamic programming tables, as even when memoization libraries are available,
they are rarely used.
Furthermore, standard implementations in imperative languages typically avoid the natural
recursive representations because of the associated costs in languages such as C++.
The efficient machine code that GHC generates for recursive functions allowed us to write
a program that faithfully reflected the mathematics, without paying great costs in run-time
efficiency.

@@

Our implementations of hidden Markov models and Viterbi's algorithm
simply duplicate 
functionality that is already implemented in the C++~package
HMMER.
These implementations form just one component of our new tool, MRFy.

>>>>

\subsection{Computing probabilities using perspicuous Haskell}

Given a hidden Markov model, 
an~established software package called HMMER (``hammer'') 
can compute the probability
that a new protein shares structure and
evolutionary history with the proteins used to train the model.
The computation finds the most likely path through the hidden Markov model.
To~make best use of floating-point arithmetic, the software computes
the \emph{logarithm} of the probability of each path, by summing
the logs of the 
probabilities on the states and edges of the path \cite{Viterbi:1967hq}.
(The logarithm function is monotonic, so the path that maximizes the
log of the probability is the most likely path.)
The~specification is shown on the left-hand side of \figref{viterbi}.
The probability $V_j^M(i)$ represents the probability of the most
likely path of the first~$i$ amino acids in the query sequence,
terminating with placement of amino acid $x_i$ in state~$M_j$.
The probabilities $V_j^D(i)$ and $V_j^D(i)$ are similar.
Equations like these can be found in more recent literature
\cite{Durbin:1998wz,Eddy:1998ut}.

@@

Another of us was curious to try to code Viterbi's equations
directly as recursive functions.
But like the classic recursive Fibonacci function, Viterbi's functions,
when implemented \naive ly,
perform extraordinarily badly.
But just like the Fibonacci function, Viterbi's functions can be
\emph{memoized}.
To~show the memoized code, we have to confess to a small
transformation of Viterbi's equations.
The logarithms on the left of \figref{viterbi} are computed when the
hidden Markov model is constructed, and the resulting numbers are
stored in a file.
Moreover, to make the file more pleasant to read, the logarithms are
negated.
(The logarithm of a probability is typically negative, and negating
the logarithm eliminates hordes of minus signs from the file.)
We~therefore implement a transformed version of Viterbi's equations,
shown on the right-hand side of \figref{viterbi}, which minimize the
negated log probability for each combination of node~$j$, amino
acid~$x_i$, and state $M_j$, $I_j$, or $D_j$.
Our implementation computes not only the minimum
negated log probability, called a \texttt{Score},
but also the path of states which leads to that most likely outcome. 
Because the cases for all three states are so similar, 
we~show only the code for the match state:
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{smallverbatim}
viterbi' :: State -> Int -> Int -> (Score, [State])
viterbi' Mat j i = ( eProb j i + minimum $ [ trans m_m Mat
                                          , trans i_m Ins
                                          , trans d_m Del 
                                          ]
                  , Mat : path )
  where (prevV, path) = viterbi prev (j-1) (i-1)
        trans t prev = prevV + tProb hmm (j-1) t
\end{smallverbatim}
% Similarly, the insertion equation is:
% \begin{verbatim}                    
%   -- consume an observation but not a node
%   | s == ins = let trans t prev =
%                    (score + tProb n t + (eProb n o), ins:path)
%                     where (score, path) = viterbi prev n (o - 1)
%                 in minimum [
%                     trans m_i mat,
%                     trans i_i ins
%                     ]
% \end{verbatim}
% Finally, the deletion equation is:
% \begin{verbatim}                    
%   -- consume a node but not an observation
%   | s == del = let trans t prev = 
%                       (score + tProb (n-1) t, del:path)
%                       where (score, path) = viterbi prev (n - 1) o
%                 in minimum [
%                     trans m_d mat,
%                     trans d_d del
%                     ]
% \end{verbatim}
Programming the algorithm in this form was not only intellectually
satisfying; keeping the code close to the math also helped us find
bugs.
The most memorable bug was that in one of our base cases, 
we~mistakenly returned an empty path of states;
the correct result was a singleton path containing the initial state.
Very quickly after observing a missing state in the output, we were
able to find the faulty case in the code.

Whatever the attractions of the code above, in our work we also
require high performance.
To make the simple recursive code efficient, we used Luke Palmer's
\texttt{Data.Memocombinators} package.
The function \verb+viterbi'+ is unmemoized,
and it makes recursive calls to \verb+viterbi+, which is memoized:
\begin{smallverbatim}
viterbi state j i = 
  Memo.memo3 (Memo.arrayRange (Mat, End)) 
             (Memo.arrayRange (0, numNodes))
             (Memo.arrayRange (0, seqlen)) 
             viterbi' state j i
\end{smallverbatim}
The value
\texttt{numNodes} represents~$\alignwidth$,
and \texttt{seqlen} represents~$N$.\remark{I hope this is right. ---NR}

This memoized recursive code performs just as well as the classic
dynamic-programming code.
And~the short fragment above is the \emph{only} part of the code
devoted to improving performance.
By contrast, standard implementations of the Viterbi algorithm, such as in HMMER,
spend much of their code 
managing dynamic-programming tables.
Haskell enabled us write simple, performant code with little effort.
%
Because the memoized version so faithfully resembles the equations in
\figref{viterbi}, we~retired the classic dynamic-programming version.

\remark{NR doesn't think the bug story is that interesting.}

%%  
%%  
%%  In this, we were grateful for the resemblance 
%%  between the mathematical description of the algorithm and the top-down 
%%  dynamic-programming approach in Haskell, which resulted in perspicuous code.
%%  
%%

\subsection{Solving new problems using perspicuous Haskell}

@@

Our implementations of hidden Markov models and Viterbi's algorithm
duplicate existing functionality.
They form just one component of our new tool, MRFy.
